---
title: "Statistical Analysis of Data"
author: Chris Malone
---

```{r source, include=FALSE, purl=FALSE}
source("setup.R")
```

```{r, purl=TRUE, echo=FALSE}
## Statistical Analysis of Data
```


------------

> ### Learning Objectives
>
> * Conducting t-tests
> * Basic model fitting 

------------


## Structure of the Vitamin Data

```{r, echo=FALSE, purl=TRUE}
### Structure of the vitamin data
```

This datasets contains information about actual intake and daily recommended intake for calories and various other vitamin.  The relationship is to be investigated across gender and the potential impact of BMI. 
Each row holds information for a single person, and the columns represent:

| Column           | Description                        |
|------------------|------------------------------------|
| RowID       | Unique row id for each person      |
| Gender           | Gender of person                   |
| BMI              | Body mass index                    |
| \<variable\>_Actual | Actual amount for \<variable\>     |
| \<variable\>_DRI   |  Daily recommended intake for \<variable\>         |

------------

> ### Download Vitamin Data to Your Local Machine
>
>  Link: [Data](https://github.com/WSU-DataScience/R-ecology-lesson/blob/gh-pages/vitamin.csv)

------------


## Data Preparation


```{r, eval=TRUE,  purl=FALSE}
vitamin <- read.csv('data/vitamin.csv', header=TRUE, sep=",")
```

Using `head()` function to view first few rows of data.

```{r, results='show', purl=FALSE}
head(vitamin)
```


The initial investigation will focus on Energy (i.e. Calorie intake).  The `select()` function from the `dplyr` package will be used to create a smaller `data.frame` from which analysis will be done.


```{r, results='show', purl=FALSE, , warning=FALSE}
library(dplyr)
vitamin_energy <- select(vitamin, RowID, Gender, BMI, Energy_Actual, Energy_DRI)
```

The resulting `vitamin_energy` data.frame from the above command.
```{r, results='show', purl=FALSE}
head(vitamin_energy)
```

> ### Goal: Compare actual intake to daily recommended intake 
>

The goal is to compare the actual intake to the daily recommended intake on an individual-by-individual level.  The `mutate()` function will be used to obtain a difference column for each individual. 

$$Difference = ( Energy\_Actual - Energy\_DRI )$$

```{r, results='show', purl=FALSE}
vitamin_energy <- vitamin_energy %>% mutate('Difference' = Energy_Actual - Energy_DRI)
```

The differences for the first few observations. 

```{r, results='show', purl=FALSE}
head(vitamin_energy)
```

> ### Understanding the Difference Variable
>
> 1. What does a negative difference mean in the context of this problem?
>
> 2. What would a difference near 0 imply?
>


## Comparing Differences

The following use of piping and the `filter()` function can be used to obtain the number of individuals that are taking in less calories than their daily recommended intake.

```{r, results='show', purl=FALSE}
vitamin_energy %>% filter(Difference < 0) %>% summarize('n' = n())
```

There are a total of $71$ observations in this `data.frame` thus, about $55/71 = 77\%$ of individuals in this dataset are not taking in enough calories relative to their daily recommended intake.

Using `ggplot()` to obtain a plot of the difference column for all individuals.

```{r, results='show', purl=FALSE, , fig.width=4, fig.align='center'}
  library(ggplot2)
  ggplot(data=vitamin_energy, aes(x="",y=Difference)) + geom_jitter(width=0.03) + geom_hline(yintercept=0) + labs(x="")
```


Getting the basic descriptive summaries using `summarize()`.
```{r, results='show', purl=FALSE}
    vitamin_energy %>% summarize('Mean Difference' = mean(Difference), 'Std'=sd(Difference), 'n'=n())
```

These summaries can be used to conduct the following statistical test.  This test will statistically determine whether or not the average difference across all WSU students is less than zero (i.e. on average are WSU students not taking in enough calories).

$$\begin{array}
 {l}
 H_{O}: \mu_{Difference} = 0 \\
 H_{A}: \mu_{Difference} < 0 \\
\end{array}$$

The test statistic for this particular statistical test is computed here.

$$\frac{(\bar{x} - \mu)}{\frac{s}{\sqrt{n}}} = \frac{(-346.1 - 0)}{\frac{486.4}{\sqrt{71}}} = -5.996$$

The apprpriate t-test can be done directly in R as follows.
```{r, results='show', purl=FALSE}
   t.test(vitamin_energy$Difference, mu=0, alternative=c('less'))
```

## Comparing Differences Across Gender

Using `ggplot()` to obtain a plot of the difference column for all individuals <strong>by</strong> Gender.

```{r, results='show', purl=FALSE}
  ggplot(data=vitamin_energy, aes(x=Gender,y=Difference, color=Gender)) + geom_boxplot() + geom_jitter(width=0.05) + geom_hline(yintercept=0)
```

Getting the basic descriptive summaries using `summarize()` <strong>by</strong> Gender.
```{r, results='show', purl=FALSE}
    vitamin_energy %>% group_by(Gender) %>% summarize('Mean Difference' = mean(Difference), 'Std'=sd(Difference), 'n'=n())
```

The statistical test here would be to compare the average deficiency in calorie intake across genders.

$$\begin{array}
 {l}
 H_{O}: \mu_{Difference:Females} =  \mu_{Difference:Males}\\
 H_{A}: \mu_{Difference:Females} \ne \mu_{Difference:Males} \\
\end{array}$$

Getting the basic descriptive summaries using `summarize()` <strong>by</strong> Gender.
```{r, results='show', purl=FALSE}
    FemaleData <- vitamin_energy %>% filter(Gender=="F") %>% select(Difference)
    MaleData <- vitamin_energy %>% filter(Gender=="M") %>% select(Difference)
    t.test(FemaleData, MaleData, alternative=c("two.sided"))
```


## Impact of BMI on Calorie Differences
Consider the following plot that allows one to understand the relationship between BMI and $(Energy\_Actual - Energy\_DRI)$ across Gender.
```{r, results='show', purl=FALSE}

  ggplot(data=vitamin_energy, aes(x=BMI, y=Difference, color=Gender)) + geom_point(aes(color=Gender)) + geom_hline(yintercept = 0) + geom_smooth(se=FALSE) + labs(y="Calorie Difference")
```


> ### Understanding Relationship between BMI and Differences
>
> 1. What general trends exist for female students?
>
> 2. What is the impact of BMI for females students?
>
> 3. Is the impact of BMI the same across Gender?


## Impact of BMI on Vitamin A

```{r, results='show', purl=FALSE}
 vitaminA_difference <- vitamin %>% mutate('Difference' = VitaminA_Actual - VitaminA_DRI) 

 ggplot(vitaminA_difference, aes(x=BMI, y=Difference, color=Gender)) + geom_point(aes(color=Gender)) + geom_hline(yintercept = 0) + geom_smooth(se=FALSE) + labs(y="VitaminA Difference")
```

## Impact of BMI on  Vitamin C

```{r, results='show', purl=FALSE}
 vitaminC_difference <- vitamin %>% mutate('Difference' = VitaminC_Actual - VitaminC_DRI) 

 ggplot(vitaminC_difference, aes(x=BMI, y=Difference, color=Gender)) + geom_point(aes(color=Gender)) + geom_hline(yintercept = 0) + geom_smooth(se=FALSE) + labs(y="VitaminC Difference")
```

## Impact of BMI on Calcium

```{r, results='show', purl=FALSE}
 Calcium_difference <- vitamin %>% mutate('Difference' = Calcium_Actual - Calcium_DRI) 

 ggplot(Calcium_difference, aes(x=BMI, y=Difference, color=Gender)) + geom_point(aes(color=Gender)) + geom_hline(yintercept = 0) + geom_smooth(se=FALSE) + labs(y="Calcium Difference")
```

## Impact of BMI on Iron

```{r, results='show', purl=FALSE}
 Iron_difference <- vitamin %>% mutate('Difference' = Iron_Actual - Iron_DRI) 

 ggplot(Iron_difference, aes(x=BMI, y=Difference, color=Gender)) + geom_point(aes(color=Gender)) + geom_hline(yintercept = 0) + geom_smooth(se=FALSE) + labs(y="Iron Difference")
```

## Example Regression Model

The smoothing function used in these plots can be changed.  For example, the following will smooth the scatterplot using a linear regression equation.
```{r, results='show', purl=FALSE}
 ggplot(Iron_difference, aes(x=BMI, y=Difference, color=Gender)) + geom_point(aes(color=Gender)) + geom_hline(yintercept = 0) + geom_smooth(method="lm",se=FALSE) + labs(y="Iron Difference")
```

The analogous linear regressio model can be fit in R as follows.  The `summary()` function provides a summary of the resulting fit.

```{r, results='show', purl=FALSE}
 lmfit <- lm(Difference ~ BMI + Gender + BMI*Gender, data=Iron_difference)
 summary(lmfit)
```





<p style="text-align: right; font-size: small;">Page build on: `r format(Sys.time())`</p>
